---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# instantiate: pre-compiled CmdStan models in R packages

<!--[![CRAN](https://www.r-pkg.org/badges/version/instantiate)](https://CRAN.R-project.org/package=instantiate)-->
[![status](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![check-cmdstanr](https://github.com/wlandau/instantiate/workflows/check-cmdstanr/badge.svg)](https://github.com/wlandau/instantiate/actions?query=workflow%3Acheck-cmdstanr)
[![check-cran](https://github.com/wlandau/instantiate/workflows/check-cran/badge.svg)](https://github.com/wlandau/instantiate/actions?query=workflow%3Acheck-cran)
[![check-fixed](https://github.com/wlandau/instantiate/workflows/check-fixed/badge.svg)](https://github.com/wlandau/instantiate/actions?query=workflow%3Acheck-fixed)
[![codecov](https://codecov.io/gh/wlandau/instantiate/branch/main/graph/badge.svg)](https://app.codecov.io/gh/wlandau/instantiate)
[![lint](https://github.com/wlandau/instantiate/workflows/lint/badge.svg)](https://github.com/wlandau/instantiate/actions?query=workflow%3Alint)

Similar to [`rstantools`](https://mc-stan.org/rstantools/) for [`rstan`](https://mc-stan.org/rstan/), the `instantiate` package builds pre-compiled [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) models into CRAN-ready statistical modeling R packages. The models compile once during installation, the executables live inside the file systems of their respective packages, and users have the full power and convenience of [`CmdStanR`](https://mc-stan.org/cmdstanr/) without any additional compilation. This approach saves time, allows R package developers to migrate from [`rstan`](https://mc-stan.org/rstan/) to the more modern [`CmdStanR`](https://mc-stan.org/cmdstanr/), and fits well with centrally maintained R installations where users may have trouble installing their own packages, diagnosing compilation errors, and setting environment variables.

# Documentation

The website at <https://wlandau.github.io/instantiate/> includes a [function reference](https://wlandau.github.io/instantiate/reference/index.html) and other documentation.

# Installation

The `instantiate` package depends on the R package [`CmdStanR`](https://mc-stan.org/cmdstanr/) and the command line tool [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan), so it is important to follow these stages in order:

1. Install the R package [`CmdStanR`](https://mc-stan.org/cmdstanr/). [`CmdStanR`](https://mc-stan.org/cmdstanr/) is not on CRAN, so the recommended way to install it is `install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))`.
2. Optional: set environment variables `CMDSTAN_INSTALL` and/or `CMDSTAN` to manage the  [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) installation. See the "Administering CmdStan" section below for details.
3. Install `instantiate` using one of the R commands below.

Type | Source | Command
---|---|---
Release | CRAN | `install.packages("instantiate")`
Development | GitHub | `remotes::install_github("wlandau/instantiate")`
Development | R-universe | `install.packages("instantiate", repos = "https://wlandau.r-universe.dev")`

## Environment variables

The `instantiate` package uses environment variables to manage the installation of [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan). An environment variable is an operating system setting with a name and a value (both text strings). In R, there are two ways to set environment variables:

1. `Sys.setenv()`, which sets environment variables temporarily for the current R session.
2. The `.Renviron` text file in you home directory, which passes environment variables to all new R sessions. the [`edit_r_environ()`](https://usethis.r-lib.org/reference/edit.html) function from the [`usethis`](https://usethis.r-lib.org/) package helps.

# Administering CmdStan

With `instantiate`, you have three different options for managing [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan). These options are called "installation types". To choose an installation type, set the `CMDSTAN_INSTALL` environment variable before installing `instantiate`. Your choice determines 

1. Whether [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) will be installed inside `instantiate` during `install.packages("instantiate")`.
2. The location where `instantiate` will automatically look for [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) at runtime.

The following table describes the role of the `CMDSTAN_INSTALL` environment variable at installation time (i.e. during `install.packages("instantiate")`).

Installation type | `CMDSTAN_INSTALL` | Install CmdStan? | Default path to CmdStan at runtime
---|---|---|---
Implicit (default) | `"implicit"` or `""` or unset | No | `cmdstanr::cmdstan_path()`, which depends on the current value of the `CMDSTAN` environment value at runtime.
Fixed | `"fixed"` | No | The value that the `CMDSTAN` environment variable originally contained at installation time, regardless of the current value of `CMDSTAN` at runtime.
Internal | `"internal"` | Yes | The internal copy of [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) inside `instantiate`.

At runtime, the `CMDSTAN_INSTALL` environment variable (and the `cmdstan_install` function argument) override `instantiate`'s automatic preference for [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan). For example, if you installed `instantiate` with `CMDSTAN_INSTALL` equal `"internal"`, you can set `Sys.setenv(CMDSTAN_INSTALL = "implicit")` at runtime to let `cmdstanr::cmdstan_path()` choose the path to [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan). You can always choose implicit or fixed installation, but you can only choose internal installation if `instantiate` was installed with an internal copy of [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan).

# Packaging Stan models

The following section explains how to create an R package with pre-compiled Stan models. This stage of the development workflow is considered "runtime" for the purposes of administering [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) as described previously.

## Structure

Begin with an R package with one or more Stan model files inside the `inst/stan/` directory. `stan_package_create()` is a convenient way to start.

```r
stan_package_create(path = "package_folder")
#> Example package named "example" created at "package_folder". Run stan_package_configure(path = "package_folder") so that the built-in Stan model will compile when the package installs.
```

At minimum the package file structure should look something like this:

```r
fs::dir_tree("package_folder")
#> package_folder
#> ├── DESCRIPTION
#> └── inst
#>     └── stan
#>         └── bernoulli.stan
```

## Configuration

Configure the package so the Stan models compile during installation. `stan_package_configure()` writes scripts `configure` and `configure.win` for this, as well as `cleanup` and `cleanup.win` to remove compiled model files. Inside each script is a call to `instantiate::stan_package_compile()` which you can manually edit to control how your models are compiled. For example, different calls to `stan_package_compile()` can compile different groups of models using different C++ compiler flags.

```r
stan_package_configure(path = "package_folder")
```

```r
fs::dir_tree("package_folder")
#> package_folder
#> ├── DESCRIPTION
#> ├── cleanup
#> ├── cleanup.win
#> ├── configure
#> ├── configure.win
#> └── inst
#>     └── stan
#>         └── bernoulli.stan
```

## Installation

Install the package just like you would any other R package. To install it from your local copy of `package_folder`, open R and run:

```r
install.packages(pkgs = "package_folder", type = "source", repos = NULL)
```

## Models

A user can now run a model from the package without any additional compilation. See the documentation of [`CmdStanR`](https://mc-stan.org/cmdstanr/index.html) to learn how to use [`CmdStanR`](https://mc-stan.org/cmdstanr/index.html) model objects.

```r
library(example)
model <- stan_package_model(name = "bernoulli", package = "example")
print(model) # CmdStanR model object
#> data {
#>   int<lower=0> N;
#>   array[N] int<lower=0,upper=1> y;
#> }
#> parameters {
#>   real<lower=0,upper=1> theta;
#> }
#> model {
#>   theta ~ beta(1,1);  // uniform prior on interval 0,1
#>   y ~ bernoulli(theta);
#> }
fit <- model$sample(
  data = list(N = 10, y = c(1, 0, 1, 0, 1, 0, 0, 0, 0, 0)),
  refresh = 0,
  iter_warmup = 2000,
  iter_sampling = 4000
)
#> Running MCMC with 4 sequential chains...
#> 
#> Chain 1 finished in 0.0 seconds.
#> Chain 2 finished in 0.0 seconds.
#> Chain 3 finished in 0.0 seconds.
#> Chain 4 finished in 0.0 seconds.
#> 
#> All 4 chains finished successfully.
#> Mean chain execution time: 0.0 seconds.
#> Total execution time: 0.6 seconds.

fit$summary()
#> # A tibble: 2 × 10
#>   variable   mean median    sd   mad     q5    q95  rhat ess_bulk ess_tail
#>   <chr>     <num>  <num> <num> <num>  <num>  <num> <num>    <num>    <num>
#> 1 lp__     -8.15  -7.87  0.725 0.317 -9.60  -7.64   1.00    7365.    8498.
#> 2 theta     0.333  0.324 0.130 0.134  0.137  0.563  1.00    6229.    7560.
```

You can write an exported user-side function in your R package to access the model. For example, you might store this code in a `R/model.R` file in the package:

```r
#' @title Fit the Bernoulli model.
#' @export
#' @family models
#' @description Fit the Bernoulli Stan model and return posterior summaries.
#' @return A data frame of posterior summaries.
#' @param y Numeric vector of Bernoulli observations (zeroes and ones).
#' @param `...` Named arguments to the `sample()` method of CmdStan model
#'   objects: <https://mc-stan.org/cmdstanr/reference/model-method-sample.html>
#' @examples
#' if (instantiate::stan_cmdstan_exists()) {
#'   run_bernoulli_model(y = c(1, 0, 1, 0, 1, 0, 0, 0, 0, 0))
#' }
run_bernoulli_model <- function(y, ...) {
  stopifnot(is.numeric(y) && all(y >= 0 & y <= 1))
  model <- stan_package_model(name = "bernoulli", package = "mypackage")
  fit <- model$sample(data = list(N = length(y), y = y), ...)
  fit$summary()
}
```

## Development

1. In your package `DESCRIPTION` file, list <https://mc-stan.org/r-packages/> in the `Additional_repositories:` field ([example in `brms`](https://github.com/paul-buerkner/brms/blob/5c09251daabd5416e3d47004cc6c62816dc53cfa/DESCRIPTION#L95-L96)). This step is only necessary while [`cmdstanr`](https://mc-stan.org/cmdstanr/) is not yet on CRAN.

```
Additional_repositories:
    https://mc-stan.org/r-packages/
```

2. In your package `DESCRIPTION` and `NAMESPACE` files, import the `instantiate` package and function `stan_package_model()`.
3. Write user-side statistical modeling functions which call the models in your package as mentioned above.
4. [`CmdStan`](https://mc-stan.org/users/interfaces/cmdstan) is too big for [CRAN](https://cran.r-project.org), so `instantiate` will not be able to access it there. So if you plan to submit your package to CRAN, please skip the appropriate code in your examples, vignettes, and tests when `instantiate::stan_cmdstan_exists()` is `FALSE`. Explicit `if()` statements like the above one in the [`roxygen2`](https://roxygen2.r-lib.org/) `@examples` work for examples and vignettes. For tests, it is convenient to use [`testthat::skip_if_not()`](https://testthat.r-lib.org/reference/skip.html), e.g. `skip_if_not(stan_cmdstan_exists())`.
5. [`pkgload::load_all()`](https://pkgload.r-lib.org/reference/load_all.html) might not compile your models. If you use [`pkgload`](https://pkgload.r-lib.org/) or [`devtools`](https://devtools.r-lib.org/) to load and develop your package, you may need to call `instantiate::stan_package_compile()` from the root directory of your package to compile your models manually.
6. For [version control](https://git-scm.com/book/en/v2/Getting-Started-About-Version-Control), it is best practice to commit only source code files and documentation. Please do not commit any compiled executable Stan model files to your repository. If you do commit them, then other users with different machines will have trouble installing your package, and your commit history will consume too much storage. For [Git](https://git-scm.com), you may add the following lines to the [`.gitigore`](https://git-scm.com/docs/gitignore) file at the root of your package: 

```
inst/stan/**
!inst/stan/**/*.*
inst/stan/**/*.exe
inst/stan/**/*.EXE
```

7. For [continuous integration](https://devguide.ropensci.org/ci.html) (e.g. on [GitHub Actions](https://github.com/r-lib/actions)), please use [`cmdstanr`](https://mc-stan.org/cmdstanr/)-based installation as explained above, and tweak your workflow YAML files as explained in that section.
8. For general information on R package development, please consult the free online book [R Packages (2e)](https://r-pkgs.org/) by [Hadley Wickham](https://github.com/hadley) and [Jennifer Bryan](https://github.com/jennybc), as well as the official manual on [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html) by the R Core Team.

# Code of Conduct

Please note that the `instantiate` project is released with a [Contributor Code of Conduct](https://github.com/wlandau/instantiate/blob/main/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

# Citation

```{r, eval = FALSE, warning = FALSE, comment = ""}
To cite package ‘instantiate’ in publications use:

  Landau WM (2023). _instantiate: A Minimal CmdStan Client for R Packages_.
  https://wlandau.github.io/instantiate/, https://github.com/wlandau/instantiate.

A BibTeX entry for LaTeX users is

  @Manual{,
    title = {instantiate: A Minimal CmdStan Client for R Packages},
    author = {William Michael Landau},
    year = {2023},
    note = {https://wlandau.github.io/instantiate/,
https://github.com/wlandau/instantiate},
  }
```
